Measurement of total hadronic differential cross sections in the LArIAT experiment

### 2018-04-30 Chapter-1/1-Theory.tex
s

\subsection{Open Questions in Neutrino Physics}
On one hand, the last three decades of experiments in neutrino oscillations brought spectacular advancements in the understanding of the oscillations pattern, measuring the neutrino mixing angles and mass splitting with a precision of less than 10\%.  On the other, they opened the field for a series of questions needing experimental answers.

>>>>

\subsection{Open Questions in Neutrino Physics}\label{ch:questions}
On one hand, the last three decades of experiments in neutrino oscillations brought spectacular advancements in the understanding of the oscillations pattern, measuring the neutrino mixing angles and mass splitting with a precision of less than 10\%.  On the other, they opened the field for a series of questions needing experimental answers.

### 2018-04-30 Chapter-2/2-LArTPCs.tex
s

\textbf{Deconvolution.} Induction and collection planes have different field responses, given the different nature of the signals on these planes: the wires on the induction planes see the inductive signal of the drifting charge, while the wires on the collection planes see the current derived from the charge entering the conductor. Thus, signals on the induction plane are bi-polar pulse and signal on the collection plane are unipolar pulses, see Figure \ref{fig:SignalProc} panel a. The first step in signal processing is deconvolution, that is a series of off-line algorithms geared towards undoing the detector effects. The result of the deconvolution step is  the production of  a comparable set waveforms on all planes presenting unipolar, approximately gaussian-like pulses (Figure \ref{fig:SignalProc} panel b). Signal from all planes are treated on equal footage beyond this point. Some LArTPC apply noise filtering in the frequency domain just after the deconvolution to clean up wire cross talk. Since signals from the LArIAT TPC are extremely clean, noise filtering is not necessary.\\

\textbf{Hit Reconstruction.} The second stage of the signal processing is the reconstruction of hits, indicating an energy deposition in the detector.  A peak finder scans the deconvolved TPC waveforms for each wire on the whole readout time looking for spikes  above  the waveform's baseline. It then fits these peaks with gaussian shapes and stores the fit parameters such as the quality of the fit, the peak time, height and area under the gaussian fit. The information resulting fron this process a single spike form a single reconstructed ``hit".
The next steps in the event reconstruction chain will then decide if rejecting hits with poor fits.
It is important to notice how the height and width of the hit depend on the topology of the event: for example, a particle running  parallel to the wire planes will leave a series of sharp hits on many consecutive wires, while a particle traveling towards the planes will leave a long, wide hit on very few wires. The height of the hits and their integral is proportional to the charge collected on the wire, so it depends on the particle type.\\

@@

\textbf{2D Clustering Reconstruction.} 
The LArIAT reconstruction of track-like objects starts by clustering hits on the collection and induction planes separately with the use of the TrajCluster clustering package\cite{Baller2016}. 
TrajCluster looks for a collection of hits in the wire-time 2D space which can be described with a line-like 2D trajectory. TrajCluster reconstructs trajectories by adding trajectory points to the leading edge of the trajectory while stepping through the 2D space of hits. Several factors determine whether a hit is added to the trajectory, including but not limited to
\begin{enumerate}
\item the goodness of the fit of the single hit,
\item the charge of the hit compared to the average charge and RMS of the hits already forming the trajectory,
\item the goodness of trajectory fit with and without the hit addition,
\item the angle between the two lines formed by the collection of hits before and after the considered hit in the trajectory.
\end{enumerate}

\textbf{3D Tracking.} The 3D tracking set of algorithms uses clusters close in time on the induction and collection planes as starting point to form a 3D track. Firstly, it construct a tentative 3D trajectory using the edges of the clusters. Then, it  projected back the tentative trajectory on to the planes and adjusts the parameters of the 3D track fit such that they minimize the distance between the fit projections and the track hits in all wire planes simultaneously.  The track algorithm can use multiple clusters in one plane, but it can never break them in smaller groups of hits. This algorithm was first developed for the ICARUS collaboration\cite{Antonello2013}.\\

\textbf{Calorimetry.} The last step in the event reconstruction chain is to assign calorimetric information to the track (or shower) objects. Calorimetry is performed separately on the different planes. A multi-step procedure is needed to retrieve the energy deposited in the TPC  from the charge seen by the wires.
For each hit associated with the track object, the calorimetry algorithms calculate the charge seen on every wire using the area underneath the gaussian fit; then they correct this raw charge by the electron life time, the electronic noise on the considered wire and the recombination effect. Lastly an overall calibration of the energy, explained in detail in section \ref{ch:energyCalibration}, is applied and the calorimetric information for the given track is assigned.
Even if calorimetry is done in 2D, it benefits from the 3D tracking information; typical information available after the calorimetric reconstruction are the total energy deposited by the particle and its stopping power $dE/dx$ at each ``track pitch", i.e. at each 2D projection on the wire plane of the 3D trajectory.

\section{The Intensity Frontier Program}
\subsection{Prospects for LArTPCs in Neutrino Physics: SBN and DUNE}
The ArgoNeut  experiment \cite{ArgoNeuT-det} initiated the US LArTPC neutrino program. Following the success of this small TPC on the NuMI beam, a wide program of LArTPCs on neutrino beams has flourished. The construction of  LArTPCs as near and far detectors at different baseline allows for the exploration of different fundamental questions in neutrino physics today.

The Short-Baseline Neutrino (SBN) \cite{Antonello:2015lea} program at Fermilab  is tasked with conclusively addressing the ``LSND and MiniBooNE anomalies" \cite{Aguilar:2001ty, Athanassopoulos:1997pv, Aguilar-Arevalo:2013pmq}, whose $\nu_e$ and $\bar\nu_e$ appearance results may be interpreted under the assumption of a new sterile neutrino.  The SBN program entails three surface LArTPCs positioned on the Booster Neutrino Beam at different distances from the neutrino production in oder to fully exploit  the L/E dependence of the oscillation pattern:  SBND (100 m from the decay pipe), MicroBooNE (450 m), and ICARUS (600 m). 
Within the oscillation context, the choice of the LArTPC technology for the SBN detectors changes the set of systematics with respect to LSND and MiniBooNE, whose detection techniques were both based on Cherenkov light.  In particular, LArTPCs provide excellent electron/photon separation \cite{Acciarri:2016sli} lacking in Cherenkov detectors which can be leveraged to abate the photon background from neutral current interactions  in $\nu_e$ searches.
MicroBooNE\cite{MicroBooNE-det}, the first detector of the SBN program to be fully operational, started its first neutrino run in October 2015. MicroBooNE is a 89 ton active volume LArTPC, single drift chamber with TPC dimensions of 2.6 m (drift) x 2.3 m (heigh) x 10.4 m (depth). MicroBooNE is positioned at a very similar L/E on the Booster neutrino beam as MiniBooNE has the scope to directly cross check the MiniBooNE oscillation measurement. 
In case MicroBooNE confirms the presence of the ``low energy excess" anomaly, SBND and ICARUS will provide the full measurement of the oscillation parameters. SBND and ICARUS are both dual drift chambers, whose active volume is respectively 112 ton and 600 ton. ICARUS is scheduled to become operational by the end of 2018 and SBND shortly after. Besides the oscillation analysis, the second main goals of SBN is to perform an extensive campaign of neutrino cross section measurements in argon. Given the importance of nuclear effects in (relatively) heavy materials, both the oscillation analysis of the SBN program and the measurements of neutrino properties in DUNE will benefit from such a campaign.

@@

%\subsection{SBN Goals}
%\subsection{Neutrino Interactions and Detection }
\subsection{Prospects for LArTPCs in GUT Physics: DUNE}
The experimental exploration of a manifestation of Grand Unified Theory is possible in DUNE thanks to its sheer mass.  In particular, proton decay searches are a capital topic of DUNE's wide non-accelerator physics program.
The key elements for a rare decay experiment are: massive active volume, long exposure, high identification efficiency and low background. 
%The limit to proton lifetime in case of absence of signal and backgrounds is set by calculating
%$$\tau/B > M\times \epsilon\times T \times 10^{32},$$ 
%where M is the detector mass in kton, $\epsilon$ the signal detection efficiency after cuts to suppress backgrounds (dependent on the considered decay mode), T is the exposure in years, B the assumed branching fraction for the considered mode and  $10^{32}$ is a factor accounting for the number of nucleons in a kton of material \cite{Bueno2007}.
Figure \ref{fig:PDKExperimentalLImit} shows the current best experimental limits on nucleon decay lifetime over branching ratio (dots). Historically, the dominant technology used in these searches has been water Cherenkov detectors: all the best experimental limits on every decay mode are indeed set by Super-Kamiokande \cite{PhysRevD.90.072005,PhysRevLett.115.121803}.  As shown in section \ref{ch:GUTsTheory}, different family of GUTs predict the proton to decay in different modes. In particular, SUSY flavored GUTs prefer the presence of kaons in the decay products, e.g. $p \rightarrow K^+ \bar{\nu}$.
It is particularly important to notice that the kaon energy for the proton decay mode $p \rightarrow K^+ \bar{\nu}$ is under Cherenkov threshold in water.  Thus, Super-Kamiokande set the limit on the lifetime for the $p \rightarrow K^+ \bar{\nu}$ mode by  relying  on photons from nuclear de-excitation and on the muon tagging in the kaon decay leptonic mode. For this reason, an attractive alternative approach to identifying nucleon decay is the use of a LArTPCs, where the kaon is directly visible in the detector. 
According to \cite{Adams:2013qkq}, DUNE will have an active volume large enough, have sufficient shielding from the surface, and will run for lengths of time sufficient to compete with Hyper-K, opening up the opportunity for the discovery of nucleon decay.

@@

In order to showcase LArIAT's utility to SBN and DUNE, we illustrate briefly two comparisons as examples: one regarding neutrino interactions and the second regarding proton decay studies.\\
The left side of figure \ref{fig:NuSimulation} shows the distribution of products in momentum spectrum and particle type as simulated in a $\nu_e$ CC interaction in DUNE (according to \cite{LeiguideOliveira:1953730}) as  a comparative example to the momentum distribution of particles in the LArIAT beamline -- shown on the right side of figure \ref{fig:NuSimulation}. The momentum spectrum in the LArIAT beamline for electrons, muons and pions -- the most abundant particles produced in  a $\nu_e$ CC interaction -- covers a wide range of the expected momentum distribution in a neutrino event.

The signature of a proton decay event in the ``LAr golden mode" is the presence of a single kaon of about 400 MeV in the detector; the momentum spectrum of the kaon pre and post FSI in such an event is shown on the left side of figure \ref{fig:PDKGENIE}. The right side of figure \ref{fig:PDKGENIE} shows the momentum spectrum of kaons in the LArIAT beamline. Kaons arriving to the LArIAT TPC are ideal for proton decay studies, since their momentum in the beamline is just above the typical momentum for kaons in a proton decay event: the majority of LArIAT kaons slow down in the TPC enough to enter the desired momentum window.

>>>>

\textbf{Deconvolution.} Induction and collection planes have different field responses, given the different nature of the signals on these planes: the wires on the induction planes see the inductive signal of the drifting charge, while the wires on the collection planes see the current derived from the charge entering the conductor. Thus, signals on the induction plane are bi-polar pulse and signal on the collection plane are unipolar pulses, see Figure \ref{fig:SignalProc} panel a). The first step in signal processing is deconvolution, that is a series of off-line algorithms geared towards undoing the detector effects. The result of the deconvolution step is  the production of  a comparable set waveforms on all planes presenting unipolar, approximately gaussian-like pulses (Figure \ref{fig:SignalProc} panel b). Signal from all planes are treated on equal footage beyond this point. Some LArTPC apply noise filtering in the frequency domain just after the deconvolution to clean up wire cross talk. Since signals from the LArIAT TPC are extremely clean, noise filtering is not necessary.\\

\textbf{Hit Reconstruction.} The second stage of the signal processing is the reconstruction of hits, indicating an energy deposition in the detector.  A peak finder scans the deconvolved TPC waveforms for each wire on the whole readout time looking for spikes  above  the waveform's baseline. It then fits these peaks with gaussian shapes and stores the fit parameters such as the quality of the fit, the peak time, height and area under the gaussian fit. The information resulting from this process on a single spike form a single reconstructed ``hit", see Figure \ref{fig:SignalProc} panel c).
The next steps in the event reconstruction chain will then decide if rejecting hits with poor fits.
It is important to notice how the height and width of the hit depend on the topology of the event: for example, a particle running  parallel to the wire planes will leave a series of sharp hits on many consecutive wires, while a particle traveling towards the planes will leave a long, wide hit on very few wires. The height of the hits and their integral is proportional to the charge collected on the wire, so it depends on the particle type.\\

@@

\textbf{2D Clustering Reconstruction.} 
The LArIAT reconstruction of track-like objects starts by clustering hits on the collection and induction planes separately with the use of the TrajCluster clustering package\cite{Baller2016}. 
TrajCluster looks for a collection of hits in the wire-time 2D space which can be described with a line-like 2D trajectory. TrajCluster reconstructs trajectories by adding trajectory points to the leading edge of the trajectory while stepping through the 2D space of hits. Several factors determine whether a hit is added to the trajectory, including but not limited to
\begin{enumerate}
\item the goodness of the fit of the single hit,
\item the charge of the hit compared to the average charge and RMS of the hits already forming the trajectory,
\item the goodness of trajectory fit with and without the hit addition,
\item the angle between the two lines formed by the collection of hits before and after the considered hit in the trajectory.
\end{enumerate}
The final product of this reconstruction stage is the collection of bidimensional clusters on each wire plane, see Figure \ref{fig:SignalProc} panel d).

\textbf{3D Tracking.} The 3D tracking set of algorithms uses clusters close in time on the induction and collection planes as starting point to form a 3D track. Firstly, it construct a tentative 3D trajectory using the edges of the clusters. Then, it  projected back the tentative trajectory on to the planes and adjusts the parameters of the 3D track fit such that they minimize the distance between the fit projections and the track hits in all wire planes simultaneously.  Tridimensional tracking can use multiple clusters in one plane, but it can never break them in smaller groups of hits. This algorithm was first developed for the ICARUS collaboration\cite{Antonello2013}. The final product of this reconstruction stage is the formation of  tridimensional objects in the TPC active volume, see Figure \ref{fig:SignalProc} panel e).\\

\textbf{Calorimetry.} The last step in the event reconstruction chain is to assign calorimetric information to the track (or shower) objects. Calorimetry is performed separately on the different planes. A multi-step procedure is needed to retrieve the energy deposited in the TPC  from the charge seen by the wires.
For each hit associated with the track object, the calorimetry algorithms calculate the charge seen on every wire using the area underneath the gaussian fit; then, they correct this raw charge by the electron life time, the electronic noise on the considered wire and the recombination effect. Lastly an overall calibration of the energy, explained in detail in section \ref{ch:energyCalibration}, is applied and the calorimetric information for the given track is assigned.
Even if calorimetry is done in 2D, it benefits from the 3D tracking information; typical information available after the calorimetric reconstruction are the total energy deposited by the particle and its stopping power $dE/dx$ at each ``track pitch", i.e. at each 2D projection on the wire plane of the 3D trajectory.

\section{The Intensity Frontier Program}
\subsection{Prospects for LArTPCs in Neutrino Physics: SBN and DUNE}
The ArgoNeut  experiment \cite{ArgoNeuT-det} initiated the US LArTPC neutrino program. Following the success of this small TPC on the NuMI beam, a wide program of LArTPCs on neutrino beams has flourished. The construction of  LArTPCs as near and far detectors at different baseline allows for the exploration of some of the fundamental questions in neutrino physics today illustrated in section \ref{ch:questions}.

The Short-Baseline Neutrino (SBN) \cite{Antonello:2015lea} program at Fermilab  is tasked with conclusively assess the nature of the ``LSND and MiniBooNE anomalies" \cite{Aguilar:2001ty, Athanassopoulos:1997pv, Aguilar-Arevalo:2013pmq}, resolving the mystery of  sterile neutrinos at the eV$^2$ scale.  The SBN program entails three surface LArTPCs positioned on the Booster Neutrino Beam at different distances from the neutrino production in oder to fully exploit  the L/E dependence of the oscillation pattern:  SBND (100 m from the decay pipe), MicroBooNE (450 m), and ICARUS (600 m). 
Within the oscillation context, the choice of the LArTPC technology for the SBN detectors changes the set of systematics with respect to LSND and MiniBooNE, whose detection techniques were both based on Cherenkov light.  In particular, LArTPCs provide excellent electron/photon separation \cite{Acciarri:2016sli} lacking in Cherenkov detectors which can be leveraged to abate the photon background from neutral current interactions  in $\nu_e$ searches.
MicroBooNE\cite{MicroBooNE-det}, the first detector of the SBN program to be fully operational, started its first neutrino run in October 2015. MicroBooNE is a 89 ton active volume LArTPC, single drift chamber with TPC dimensions of 2.6 m (drift) x 2.3 m (heigh) x 10.4 m (depth). MicroBooNE is positioned at a very similar L/E on the Booster neutrino beam as MiniBooNE has the scope to directly cross check the MiniBooNE oscillation measurement. 
In case MicroBooNE confirms the presence of the ``low energy excess" anomaly, SBND and ICARUS will provide the full measurement of the oscillation parameters. SBND and ICARUS are both dual drift chambers, whose active volume is respectively 112 ton and 600 ton. ICARUS is scheduled to become operational by the end of 2018 and SBND shortly after. Besides the oscillation analysis, the second main goals of SBN is to perform an extensive campaign of neutrino cross section measurements in argon. Given the importance of nuclear effects in (relatively) heavy materials, as discussed in section \ref{ch:NuInt}, both the oscillation analysis of the SBN program and the measurements of neutrino properties in DUNE will benefit from such a campaign.

@@

\subsection{Prospects for LArTPCs in GUT Physics: DUNE}
The experimental exploration of a manifestation of Grand Unified Theory is possible in DUNE thanks to its sheer mass.  In particular, proton decay searches are a capital topic of DUNE's wide non-accelerator physics program.
The key elements for a rare decay experiment are: massive active volume, long exposure, high identification efficiency and low background. 
%The limit to proton lifetime in case of absence of signal and backgrounds is set by calculating
%$$\tau/B > M\times \epsilon\times T \times 10^{32},$$ 
%where M is the detector mass in kton, $\epsilon$ the signal detection efficiency after cuts to suppress backgrounds (dependent on the considered decay mode), T is the exposure in years, B the assumed branching fraction for the considered mode and  $10^{32}$ is a factor accounting for the number of nucleons in a kton of material \cite{Bueno2007}.
Figure \ref{fig:PDKExperimentalLImit} shows the current best experimental limits on nucleon decay lifetime over branching ratio (dots). Historically, the dominant technology used in these searches has been water Cherenkov detectors: all the best experimental limits on every decay mode are indeed set by Super-Kamiokande \cite{PhysRevD.90.072005,PhysRevLett.115.121803}.  As shown in section \ref{ch:GUTsTheory}, different family of GUTs predict the proton to decay in different modes. In particular, SUSY flavored GUTs prefer the presence of kaons in the decay products, e.g. $p \rightarrow K^+ \bar{\nu}$.
It is particularly important to notice that the kaon energy for the proton decay mode $p \rightarrow K^+ \bar{\nu}$ is under Cherenkov threshold in water.  Thus, Super-Kamiokande set the limit on the lifetime for the $p \rightarrow K^+ \bar{\nu}$ mode by  relying  on photons from nuclear de-excitation and on the muon tagging in the kaon decay leptonic mode. For this reason, an attractive alternative approach to identifying nucleon decay is the use of a LArTPCs, where the kaon is directly visible in the detector. 
According to \cite{Adams:2013qkq}, DUNE will have an active volume large enough, have sufficient shielding from the surface, and will run for lengths of time sufficient to compete with Hyper-K, opening up the opportunity for the discovery of nucleon decay.

@@

In order to showcase LArIAT's utility to SBN and DUNE, we illustrate briefly two comparisons as examples: one regarding neutrino interactions and the second regarding proton decay studies.\\
The left side of figure \ref{fig:NuSimulation} shows the distribution of products in momentum spectrum and particle type as simulated in a $\nu_e$ CC interaction in DUNE (according to \cite{LeiguideOliveira:1953730}); the range of these distribution is to compare with  the momentum distribution of light particles in the LArIAT beamline -- shown on the right side of figure \ref{fig:NuSimulation}. The momentum spectrum in the LArIAT beamline for electrons, muons and pions -- the most abundant particles produced in  a $\nu_e$ CC interaction -- covers a wide range of the expected momentum distribution in a neutrino event.

The signature of a proton decay event in the ``LAr golden mode" is the presence of a single kaon of about 400 MeV in the detector; the momentum spectrum of the kaon pre and post FSI in such an event as simulated by GENIE is shown on the left side of figure \ref{fig:PDKGENIE}. The right side of figure \ref{fig:PDKGENIE} shows the momentum spectrum of kaons in the LArIAT beamline. Kaons arriving to the LArIAT TPC are ideal for proton decay studies, since their momentum in the beamline is just above the typical momentum for kaons in a proton decay event: the majority of LArIAT kaons slow down in the TPC enough to enter the desired momentum window.
